Transmembrane helices are also an overlooked source of major histocompatibility complex class II epitopes and evolutionary more conserved than expected by chance

### 2021-02-17 bbbq_article.tex
Process feedback @bogeert. Thanks Geert! Progress #179
Moved Introduction part to Discussion
Process feedback until 'Measuring TMH epitopes', progress #179

% GvdB:
% The text below here is more discussion 
% as it is already interpretation of the findings. 
% I would add one short paragraph, 
% where you reiterate the most important findings:
% In this study, we show that ...(and than basically 
% the abstract in slightly different words and slightly 
% more detail). End with a statement of why it is 
% important (vaccin development based on TMHs?)...

% RJCB: This was suggested for removal, until the Methods:
% \paragraph{Does MHC-I present TMH-derived epitopes from pathogens as often?}
% 
% It is important that MHC-I presents both the peptides of the 
% (healthy) cell, as well as possible pathogen-derived peptides when the
% cell is infected. As described above, MHC-I presents TMH-derived epitopes 
% from the human host more often. It is unknown if MHC-I has the same
% dedication to present epitopes that stem from TMHs derived from
% proteins produced by pathogen, either would the pathogen
% be a virus or 
% a bacterium.
% 
% \paragraph{MHC-II is expected to present TMHs}
% 
% If presentation of TMHs on MHC-I would bring an evolutionary advantage 
% in the recognition of pathogens by the immune system, 
% it would follow that this is equally important for MHC-II, 
% especially as the help of CD4+ T cells is needed for a long lasting CD8+ T cell 
% response (\cite{novy2007cd4}). 
% The mechanism to detect foreign TMHs by MHC-II would be unknown, 
% similar to the discovery that MHC-I presents TMHs.
% 
% \paragraph{Does MHC-II present TMH-derived epitopes from pathogens as often?}
% 
% It is unknown if MHC-II, like MHC-I, 
% presents TMH-derived epitopes as often, either 
% in humans,
% bacteria or viruses.
% 
% \paragraph{Selection undetectable in whole proteome}
% 
% In general, on would hope that evolutionary selection results in
% an immune system that as most attentive for loci that are
% essential for a virus, as these will be most conserved.
% In SARS-CoV-2, for example, there is preliminary evidence that the strongest
% selection pressure is upon residues that changes its 
% virulence (\cite{velazquez2020positive}).
% These loci, however, only account for a small part of a pathogen's proteome.
% Additionally, these essential parts differ widely between pathogens.
% Because of this scarcity and variance in targets, 
% one can imagine that the human immune system 
% is not tailored to detect these sites, 
% as hinted by upon by the aforementioned influenza study.
% 
% \paragraph{Selection may be detectable in TMHs}
% 
% TMHs, on the other hand, also have their function constraints, 
% yet can occur multiple time a pathogen's proteome.
% One can safely assume a pathogen's proteome contains multiple TMHs.
% Therefore, it may be beneficial for the host
% if its immune system would be more attentive towards TMHs.
% And maybe this has already happened: MHC-I already detects hydrophobic
% peptides. This feature, however, may also be caused by selection
% to detect hydrophobic regions in the soluble proteins of pathogens.
% It is unknown, when focusing on TMHs only, if a signal of selection
% can be detected.
% 
% \paragraph{Use of protein data in phylogenetics}
% 
% When using DNA sequences, one can use a skewed rate
% in non- versus synonymous mutation
% to detect the signature of selection (\cite{murrell2015gene}).
% Using AA sequences, however, has its advantages,
% as its is closer to the actual phenotype
% selection acts upon: DNA may never be transcribed to RNA,
% or its RNA may never be translated (\cite{diz2012proteomics}).
% When using a proteome in phylogenetic research,
% we know that the majority of proteins are selected to just
% maintain their function most of the time, where
% is the time spans there is selection, only a few AAs
% can actually increase the 'fitness' of the
% protein (\cite{anisimova2009investigating}).
% There, when generalizing the dynamics of mostly purifying selection (to maintain
% a protein's function) and a short duration of positive selection,
% those genes that are selected cannot be detected (\cite{yang2000statistical}).

@@

To determine the percentages of epitopes overlapping
with TMHs, we use reference proteomes of 
a human, viral and bacterial nature.
An overview of Uniprot IDs is shown in table \ref{tab:uniprot_ids}.
Instead of using the full human reference proteome, we
use the representative human reference proteome 
as used previously (\cite{bianchi2017}).

@@

To determine if epitopes derived from TMHs are presented at all,
we started from epitopes sequenced from elution studies
for MHC-I (\cite{schellens2015comprehensive}) 
and MHC-II (\cite{bergseng2015different}).
For each epitope, the human representative reference proteome
was searched for its origin.
We kept only the epitopes of which were found exactly once
on the proteome.
We predicted the topology of the proteome
and counted how often the epitopes overlapped
with a predicted TMH.
In this analysis, both TMHMM and 
PureseqTM (see section 'Prediction software used') 
were used to predict the topology.
The full analysis can be found
at \url{https://github.com/richelbilderbeek/bbbq_article_issue_157}.

@@

To determine the percentages of MHC-I and MHC-II epitopes overlap
with TMHs, we used mostly the same analysis as described in \cite{bianchi2017}.
To summarize: from a proteome all possible 9-mers are derived. For each
of these peptides, it is measured if it has been part of a 
TMH (at least overlapping 1 residue), 
as well as if the peptide is a binder to an MHC-I haplotype.
For MHC-II, 14-mers are used, as these are the most frequently occurring
epitope length (\cite{bergseng2015different}).

This study differs in some aspects, described below in more detail,
which is the definition of what a binder is,
the addition of using MHC-II haplotypes, the software used,
and the significance level to determine if TMH-derived peptides are as 
likely to be a binder as expected by chance.
These deviations are either a refinement of the previous method or
a pragmatic choice made due to the extension of the original experiment.
Additionally, instead of only using a human proteome, in this study
also a viral and bacterial proteome are analyzed.

The definition of a binder differs from \cite{bianchi2017}:
in this study a peptide is called a binder if, within a haplotype, 
any of its 9-mer peptides have an IC50 value in the lowest 2\% of 
the epitopes within a 
\emph{proteome} (see tables \ref{tab:tmh_binders_mhc1} and \ref{tab:tmh_binders_mhc2}
for values), whereas the original study defined
a binder as having an IC50 in the lowest 2\% 
of the peptides within a \emph{protein}.
% See https://github.com/richelbilderbeek/bianchi_et_al_2017/blob/72e6755a31d400158368509fd80a41e984677ab1/predict-binders.R#L17

The 13 MHC-I haplotypes used in this study are the same as 
the ones in \cite{bianchi2017}.
The MHC-II haplotypes used additionally are selected 
to occur with a phenotypic frequency of at least 14\% in
the human population (\cite{greenbaum2011functional}),
resulting in 21 haplotypes.
When using an MHC-II haplotype, instead of using 9-mers, 14-mers are
used, as these are the most common MHC-II epitope size,
as found in the elution study (\cite{bergseng2015different}).

@@

>>>>

% GvdB:
% Add one short paragraph, 
% where you reiterate the most important findings:
% In this study, we show that ...(and than basically 
% the abstract in slightly different words and slightly 
% more detail). End with a statement of why it is 
% important (vaccin development based on TMHs?)...

@@

To determine the percentages of epitopes overlapping
with TMHs, we used reference proteomes of 
a human, viral and bacterial nature.
An overview of Uniprot IDs is shown in table \ref{tab:uniprot_ids}.
Instead of using the full human reference proteome, we
used the representative human reference proteome 
as used previously (\cite{bianchi2017}).

@@

To determine if epitopes derived from TMHs are presented at all,
we started from epitopes identified in elution studies
for MHC-I (\cite{schellens2015comprehensive}) 
and MHC-II (\cite{bergseng2015different}).
For each of these peptides, the human representative reference proteome
was searched for its origin.
We kept only the epitopes which were uniquely present
in the proteome.
We predicted the topology of the proteome
and counted how often the epitopes overlapped
with a predicted TMH.
In this analysis, both TMHMM and 
PureseqTM (see section 'Prediction software used') 
were used to predict the topology.
The full analysis can be found
at \url{https://github.com/richelbilderbeek/bbbq_article_issue_157}.

@@

To determine the percentages of MHC-I and MHC-II epitopes overlapping
with TMHs, we used mostly the same analysis as described in \cite{bianchi2017}.
To summarize: from a proteome all possible 9-mers are derived. For each
of these peptides, it was determined if it was part of a 
TMH (at least overlapping with 1 residue), 
and if the peptide bound to an MHC-I haplotype.
For MHC-II, 14-mers were used, as these are the most frequently occurring
epitope length (\cite{bergseng2015different}).

This study differs in some aspects from our previous work (\cite{bianchi2017}), 
described below in more detail.
The main differences are our definition of what a so-called binder is, 
the inclusion of both MHC-I and MHC-II haplotypes, 
the prediction software used, 
and the significance level to determine if TMH-derived peptides will bind.
These deviations are either a refinement of our previous approach or
a pragmatic choice made due to the extension of the original experiment.
Additionally, instead of only using a human proteome, this study
also includes a viral and bacterial proteome.

The definition of a binder differs from \cite{bianchi2017}:
in our current study a peptide was called a binder if, within a certain haplotype, 
any of its 9-mer peptides have an IC50 value in the lowest 2\% of 
the epitopes within a 
\emph{proteome} (see tables \ref{tab:tmh_binders_mhc1} and \ref{tab:tmh_binders_mhc2}
for values), whereas the original study defined
a binder as having an IC50 in the lowest 2\% 
of the peptides within a \emph{protein}.
% See https://github.com/richelbilderbeek/bianchi_et_al_2017/blob/72e6755a31d400158368509fd80a41e984677ab1/predict-binders.R#L17
We believe our revised definition precludes bias of proteins 
that give rise to no or only very few MHC epitopes

The 13 MHC-I haplotypes used in this study are the same as 
we used in our previous study (\cite{bianchi2017}).
The MHC-II haplotypes used additionally were selected 
to occur with a phenotypic frequency of at least 14\% in
the human population (\cite{greenbaum2011functional}),
resulting in 21 haplotypes.
When using an MHC-II haplotype, instead of using 9-mers, 14-mers were
used, as these are the most common MHC-II epitope size,
as found in the elution study (\cite{bergseng2015different}).

@@

% RJCB: moved this from Introduction to Discussion.
%
% \paragraph{Does MHC-I present TMH-derived epitopes from pathogens as often?}
% 
% It is important that MHC-I presents both the peptides of the 
% (healthy) cell, as well as possible pathogen-derived peptides when the
% cell is infected. As described above, MHC-I presents TMH-derived epitopes 
% from the human host more often. It is unknown if MHC-I has the same
% dedication to present epitopes that stem from TMHs derived from
% proteins produced by pathogen, either would the pathogen
% be a virus or 
% a bacterium.
% 
% \paragraph{MHC-II is expected to present TMHs}
% 
% If presentation of TMHs on MHC-I would bring an evolutionary advantage 
% in the recognition of pathogens by the immune system, 
% it would follow that this is equally important for MHC-II, 
% especially as the help of CD4+ T cells is needed for a long lasting CD8+ T cell 
% response (\cite{novy2007cd4}). 
% The mechanism to detect foreign TMHs by MHC-II would be unknown, 
% similar to the discovery that MHC-I presents TMHs.
% 
% \paragraph{Does MHC-II present TMH-derived epitopes from pathogens as often?}
% 
% It is unknown if MHC-II, like MHC-I, 
% presents TMH-derived epitopes as often, either 
% in humans,
% bacteria or viruses.
% 
% \paragraph{Selection undetectable in whole proteome}
% 
% In general, on would hope that evolutionary selection results in
% an immune system that as most attentive for loci that are
% essential for a virus, as these will be most conserved.
% In SARS-CoV-2, for example, there is preliminary evidence that the strongest
% selection pressure is upon residues that changes its 
% virulence (\cite{velazquez2020positive}).
% These loci, however, only account for a small part of a pathogen's proteome.
% Additionally, these essential parts differ widely between pathogens.
% Because of this scarcity and variance in targets, 
% one can imagine that the human immune system 
% is not tailored to detect these sites, 
% as hinted by upon by the aforementioned influenza study.
% 
% \paragraph{Selection may be detectable in TMHs}
% 
% TMHs, on the other hand, also have their function constraints, 
% yet can occur multiple time a pathogen's proteome.
% One can safely assume a pathogen's proteome contains multiple TMHs.
% Therefore, it may be beneficial for the host
% if its immune system would be more attentive towards TMHs.
% And maybe this has already happened: MHC-I already detects hydrophobic
% peptides. This feature, however, may also be caused by selection
% to detect hydrophobic regions in the soluble proteins of pathogens.
% It is unknown, when focusing on TMHs only, if a signal of selection
% can be detected.
% 
% \paragraph{Use of protein data in phylogenetics}
% 
% When using DNA sequences, one can use a skewed rate
% in non- versus synonymous mutation
% to detect the signature of selection (\cite{murrell2015gene}).
% Using AA sequences, however, has its advantages,
% as its is closer to the actual phenotype
% selection acts upon: DNA may never be transcribed to RNA,
% or its RNA may never be translated (\cite{diz2012proteomics}).
% When using a proteome in phylogenetic research,
% we know that the majority of proteins are selected to just
% maintain their function most of the time, where
% is the time spans there is selection, only a few AAs
% can actually increase the 'fitness' of the
% protein (\cite{anisimova2009investigating}).
% There, when generalizing the dynamics of mostly purifying selection (to maintain
% a protein's function) and a short duration of positive selection,
% those genes that are selected cannot be detected (\cite{yang2000statistical}).
